---
title: "Week 5 lab BST 210"
author: "Daniel Herrera, Willow Duffell, Lauren Mock"
date: "10/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Group 4 Members:

Daniel Herrera  
Willow Duffell  
Lauren Mock

## Question 1

**General area/domain/subject area**

The general subject area of this project is bone marrow transplant outcomes in children with hematologic diseases. 

## Question 2

**Dataset and source**

We will use data obtained from the UCI Machine Learning Repository: Center for Machine Learning and Intelligent Systems that was originally collected by researchers in the Faculty of Automatic Control, Electronics and Computer Science at Silesian University of Technology in Poland. This data set includes children with hematologic diseases, such as leukemia, who underwent bone marrow transplants. It includes information on the donor, the recipient, and the recipient’s response to the treatment. 

The source of this data set can be found here: 
https://archive.ics.uci.edu/ml/datasets/Bone+marrow+transplant%3A+children

And a downloadable version of the data set can be found here: https://www.kaggle.com/adamgudys/bone-marrow-transplant-children

## Question 3

**Primary questions**

Our primary goals are to predict (before transplantation) whether or not a patient will survive, and to understand which factors are the most important in determining survival. We are particularly interested in determining the importance of the closeness of the match between the recipient and the donor for survival. 

## Question 4

**Secondary questions**

There is a lot of controversy around the source of hematopoietic stem cells used in transplantation (either peripheral blood (PBSC) or bone marrow (BM)). It would be interesting to see if there is an association between these two stem-cell sources and the likelihood of survival, and we will definitely need to be careful with confounding when addressing this question.

## Question 5

**Outcome(s)/endpoint(s)**

Our primary outcome is survival status at the end of the follow-up period, which is a binary variable. We are interested in predicting a patient’s probability of survival, as well as understanding the relative importance of factors that influence survival. 

## Question 6

**Draft Statistical Analysis Plan**

Draft Statistical Analysis Plan: 

*I. Explain important variables (using "domain” knowledge)*

  a. Identify potential confounders and effect measure modifiers (EMM) 

  b. Will require an understanding of how our variables are associated 

*II. Exploratory Data Analysis (EDA using above)* 

  a. Visualizations (scatterplots - with jitter since survival 0-1, correlation matrices) 

  b. Visualize confounding and EMM Summary statistics 

*III. Candidate Variables* 

  a .Choose best variables to use to predict survival 

*IV. Model Selection*

a. Primary Question 

  Prediction of survival 

   1) Goal is to predict survival at end of follow-up (binary variable) 

   2) Does not need to account for confounding, because we only want the best possible prediction of survival 

   3) Perform logistic regression (binary outcome) with all variables that may be important  
        - Trim model - remove unnecessary variables to avoid unnecessary complexity/overfitting 
        - Backwards removal using adjusted R^2 (since we are more concerned with maximizing our         predictive ability--the amount of variability in y explained by our covariates) 

   4) Explore K-means as a non-parametric method of predicting survival 

  Determining Risk Factors for Mortality (Association) 

   1) Goal is to determine the most important risk factors in determining whether or not a patient survives 
        - This will only include data that is collected before transplantation 

   2) Must account for confounding, because we care about the strength of the relationship between each variable and survival 

   3) Perform logistic regression with all variables that may be important 
        - Trim model - remove unnecessary variables to avoid unnecessary complexity/overfitting
        - Backwards removal using adjusted R^2 (since we are more concerned with maximizing our predictive ability--the amount of variability in y explained by our covariates) 

b. Secondary Question--Source of hematopoietic stem cells 

   1) Goal is to determine association between the two stem-cell source methods (peripheral blood and bone marrow) and their outcome of survival, presumably using the relevant covariates discovered in our model above 

   2) Confounders will be important here, since we don’t yet know which type of patients receive which treatment 

   3) Using Association model results from our primary question, add variable stem_cell_source and run an ANOVA test to determine which source has a better outcome of survival  

*V. Check for multicollinearity using variance inflation factors* 

a. We expect many of our variables to be highly correlated 

*VI. Validate model using Cross-Validation, where appropriate*

a. Samples of the training/testing data 

b. How well would the prediction model perform on other data? 

*VII. Run sensitivity analysis using Ridge, LASSO, or Elastic Net*

## Question 7

**Biggest challenges foreseen**

One of the main challenges we foresee in answering our proposed questions is our lack of subject matter knowledge. In addressing which risk factors are most important in determining survival, we will need to have a comprehensive understanding of the potential confounders in our data set to ensure that the coefficients in our models are meaningful. In our secondary analysis, we will need to understand why a patient may be selected for a bone marrow transplant vs. a peripheral blood transplant, which will require a domain expert. We will also need to account for the many variable types included in our data, including numeric, categorical, and indicator.

## Question 8

**Domain expertise sought**

We will need to find a domain expert who has experience with hematologic diseases or bone marrow transplants in order to better understand our dataset and its variables. This will be especially important in identifying potential confounders, which cannot be identified through any statistical process alone. This will also give us some insight into possible effect measure modifiers. For now, we have discussed our dataset with Daniel’s fiancee, who is a doctor, and Willow’s mom, who is a nurse practitioner. Moving forward, we will be seeking additional expertise. 

## Question 9

**What software package(s) will you use to complete this project?**

We will all be using R.

## Question 10

**Complete an initial round of exploratory analyses on your data that would be relevant to your plan and responses above, and include any plots, summaries, code and output.  Please include exploratory analysis for outcome(s) of continuous form however/wherever possible even if your ultimate goals/questions involve a different form of outcome data such as binary, polytomous, etc. (You may consider this initial analysis as a potential sub-analysis later on.)**

```{r, warning=FALSE, include = FALSE}
library(tidyverse)
library(pander)
library(ggthemes)
library(gtools)
library(splitstackshape)
library(ResourceSelection)
library(LogisticDx)
library(grid)
library(gridExtra)

bone <- read.csv("data.csv")
```


### Explore Missing Values

Before we begin with any visualizations and summary statistics, I would like to explore  missing-ness within our data set. 

```{r}

# lets look at a table of all missing values
colSums(is.na(bone))
```

Great, no missing data (spoiler: there is). Oh wait, there are "?" marks that we should examine. We see that there are actually a bit of "?" in our data set, so this may need to be addressed later, particularly for variables such as recipient CMV (14 ?'s), CMV_status (16) and extensive_chronic_GvHD (31 ?'s). For now, we will replace the "?"s in the data set.

```{r}
# let us see the number of ?'s that appear
sort(colSums(bone == "?"), decreasing = TRUE) [1:10]


# replace ?'s with true NA's
bone <- mutate_all(bone, ~replace(., . == "?", NA)) 

# Now we can see all of the ?'s have been replaced with NA values that are picked up
sort(colSums(is.na(bone)), decreasing = TRUE) [1:10]
```

Well, it turns out there are some other problematic values. Some time-to-event variables have entries of 1,000,000, which will be challenging to handle in our EDA.

To illustrate, let us see the number of occurrences of the value 1000000. This value frequently occurs in the data amongst three variables. It seems as though these large values indicate that the event never occurred for that subject, but we will need to confirm this. If our assumption is correct, then we will need to seek out advice on how to proceed. 

For now, we will leave these values as is. We don't want to replace them with NA, because they do not appear to be missing values.

```{r}
# let us see the number of ?'s that appear
sort(colSums(bone == "1000000"), decreasing = TRUE) [1:3]

# replace with NA
# bone <- mutate_all(bone, ~replace(., . == "1000000", NA)) 

```

Let's also make sure our true missing values do not exceed 5% of each column.

```{r}

# look of missing values as % of column
missing_vals <- bone %>% 
  is.na() %>%
  colSums() %>%
  `/`(nrow(bone)) %>%
  `*`(100)

missing_vals[missing_vals >= 5] %>% names()

```

Only these three columns have > 5% missing values. We will need to investigate these missing values later if we choose to use these variables in our analysis.


### Basic Data Exploration

Now we can explore some more standard EDA. Lets familiarize ourselves with the data set by briefly viewing all the columns, column types, a brief amount of data. We can then explore summary statistics. Note: we will have to change some of our data from character data to numeric data. 

#### Numerical Variables: 
donor_age, recipient_age, recipient_body_mass, CD34_x1e6_per_kg...CD34kgx10d6, CD3_x1e8_per_kg, CD3_to_CD34_ratio, ANC_recovery, PLT_recovery, time_to_acute_GvHD_III_IV, survival_time

#### Categorical Variables: 
donor_ABO, recipient_age_int, recipient_ABO, disease, CMV_status, HLA_match..out.of.10., antigen, allele, HLA_group_1 

#### Binary Variables:
donor_age_below_35 (yes/no), donor_CMV (present/absent), recipient_age_below_10 (yes/no), recipient_rh (plus/minus), recipient_CMV (present/absent), disease_group (malignant/nonmalignant), gender_match(female_to_male/other), ABO_match (matched/mismatched), HLA_mismatch (matched, mismatched), risk_group (high/low), stem_cell_source (peripheral_blood/bone_marrow),tx_post_relapse (yes/no), acute_GvHD_II_III_IV (yes/no), acute_GvHD_III_IV (yes/no), extensive_chronic_GvHD (yes/no), relapse (yes/no), survival_status (yes,no)

```{r}
glimpse(bone)
```

### Explore donor data

```{r}

summary(bone$donor_age)

# plot histogram of age distribution
ggplot(bone) + 
  geom_histogram(aes(x = donor_age, y = ..density..), fill = "#990000", col = "black", binwidth = 1) + 
  ggtitle("Age of donors", ) +
  xlab("Age") +
  ylab("Density") +
  theme_wsj()+ theme(axis.title=element_text(size=12))

# summary stats for donor blood type
bone %>% 
  group_by(donor_ABO) %>% 
  summarise(n = length(donor_ABO),
            proportion = length(donor_ABO)/nrow(bone))

# summary stats for donor cytomegalovirus status
bone %>% 
  group_by(donor_CMV) %>% 
  summarise(n = length(donor_CMV),
            proportion = length(donor_CMV)/nrow(bone))

```

So far we can see the following:

1) most of our donors are between 20 and 50 years old
2) donors' blood types are as follows: type O (39.0%), type A (38.0%), type AB (8.0%), B (15.0%)
3) donors' cmv status follows: absent (60.4%), present (38.5%), missing (0.1%)



### Explore recipient data

```{r}

summary(bone$recipient_age)

# plot histogram of age distribution
bone %>% 
ggplot() + 
  geom_histogram(aes(x = recipient_age, y=..density..), fill = "#990000", col = "black", binwidth = 1) + 
  ggtitle("Age of recipients" ) +
  xlab("Age") +
  ylab("Density") +
  theme_wsj()+ theme(axis.title=element_text(size=12))

# summary stats for blood type
bone %>% 
  group_by(recipient_ABO) %>% 
  summarise(n = length(recipient_ABO),
            proportion = length(recipient_ABO)/nrow(bone))

# summary stats for recipient cytomegalovirus status
bone %>% 
  group_by(recipient_CMV) %>% 
  summarise(n = length(recipient_CMV),
            proportion = length(recipient_CMV)/nrow(bone))

```

So far we can see the following:

1) our recipients are mostly between 5 and 15 years old (min age is 0.6 years old and max age is 20 years old)
2) recipients' blood types are as follows: type O (25.7%), type A (40.1%), type AB (7.0%), B (26.7%), NA (0.5%)
3) recipients' cmv status follows: absent (39.0%), present (53.5%), missing (7.5%)



Now we can try to have a look at recipients' weight, rh factor (+ or -), and risk group

Note: Changed data type of recipient body mass from character to numeric
(Assuming weight has units of lbs)

```{r}
# plot histogram of body mass distribution
bone %>% 
ggplot() + 
  geom_histogram(aes(x = as.numeric(recipient_body_mass), y = ..density..),
                 fill = "#990000", col = "black", binwidth = 5) + 
  ggtitle("Weight of Recipients", ) +
  scale_x_continuous(breaks = seq(0, 110, by =10)) +
  xlab("Weight (lbs)") +
  ylab("Density") +
  theme_wsj()+ theme(axis.title=element_text(size=12))

# summary stats for recipient rh factor
bone %>% 
  group_by(recipient_rh) %>% 
  summarise(n = length(recipient_rh),
            proportion = length(recipient_rh)/nrow(bone))

# summary stats for recipient risk group
bone %>% 
  group_by(risk_group) %>% 
  summarise(n = length(risk_group),
            proportion = length(risk_group)/nrow(bone))
```

So far we can see the following about our recipients:

1) our recipients weighed between 15lbs to 60lbs (min weight is 6 lbs and max weight is 103.4 lbs)
2) recipients' rh factor are as follows: plus (84.5%), minus (14.4%), NA (1.1%)
3) risk status follows: high (36.9%), low (63.1%)


### Scatterplots

Looking only at the numerical variables next:

```{r}
bone_num <- bone %>% select(c(donor_age, recipient_age, recipient_body_mass, CD34_x1e6_per_kg...CD34kgx10d6, CD3_x1e8_per_kg, CD3_to_CD34_ratio, ANC_recovery, PLT_recovery, time_to_acute_GvHD_III_IV, survival_time, survival_status))

bone_num <- sapply(bone_num, as.numeric)
```

```{r, fig.height=10, fig.width=10}
pairs(bone_num)
```

We see at first glance based off of the correlation scatter plots that recipient age and body mass are very highly correlated (as we would expect). We also see many plots with binary variables, and several plots with points that are very far from the main cloud of data.


We can also use a correlation matrix to see the relationships between our variables, where applicable. The data is quite large with many binary variables, so perhaps our visualizations will aid us more. 

```{r}
# correlation matrix
cor(bone_num, use= "complete.obs") %>% as.data.frame() %>% pander()
```



### Compare outcome variable (survival_status) and numeric variables

1. Explore our binary outcome variable

In order to explore our outcome variable of interest, survival status after bone marrow transplant, we must first change it to numeric. This graph shows us that 45.5% of our bone marrow transplant patients unfortunately did not survive, while 54.5% did survive. 

```{r, warning=FALSE}

bone$survival_status <- ifelse(bone$survival_status == 1, "Died", "Survived") %>% as.factor()

bone %>% 
ggplot(aes(x = survival_status)) + 
  geom_histogram(fill = "#990000", col = "black", binwidth = 5, stat = "count") + 
  ggtitle("Survival Status", ) +
  xlab("Survival") +
  ylab("Count") +
  theme_wsj()+ theme(axis.title=element_text(size=12))


bone %>% 
  group_by(survival_status) %>% 
  summarise(n = length(survival_status),
            proportion = length(survival_status)/nrow(bone))
```


2. Look at binary outcome with continuous predictors

donor_age, recipient_age, recipient_body_mass, CD34_x1e6_per_kg...CD34kgx10d6, CD3_x1e8_per_kg, CD3_to_CD34_ratio, ANC_recovery, PLT_recovery, time_to_acute_GvHD_III_IV, survival_time

```{r, warning = FALSE, fig.height=12, fig.width=10}

# construct plots of continuous predictors 
c1 <- bone %>% 
  ggplot(aes(x = survival_status, y = donor_age)) +
  geom_boxplot(fill = "#990000", col = "black") +
  ggtitle("Age of Donor") +
  xlab("Survival Status") +
  ylab("Age of Donor") +
  theme_wsj()+ theme(axis.title=element_text(size=12),
                     plot.title = element_text(size = 20))


c2 <- bone %>% 
  ggplot(aes(x = survival_status, y = recipient_age)) +
  geom_boxplot(fill = "#990000", col = "black") +
  ggtitle("Age of Recipient") +
  xlab("Survival Status") +
  ylab("Age of Recipient") +
  theme_wsj()+ theme(axis.title=element_text(size=12),
                     plot.title = element_text(size = 20))


c3 <- bone %>% 
  ggplot(aes(x = survival_status, y = as.numeric(recipient_body_mass))) +
  geom_boxplot(fill = "#990000", col = "black") +
  ggtitle("Recipient Body Mass") +
  xlab("Survival Status") +
  ylab("Recipient Body Mass") +
  theme_wsj()+ theme(axis.title=element_text(size=12),
                     plot.title = element_text(size = 15))


c4 <- bone %>% 
  ggplot(aes(x = survival_status, y = as.numeric(CD34_x1e6_per_kg...CD34kgx10d6))) +
  geom_boxplot(fill = "#990000", col = "black") +
  ggtitle("CD34+ Dosage") +
  xlab("Survival Status") +
  ylab("CD34+ Dosage") +
  theme_wsj()+ theme(axis.title=element_text(size=12),
                     plot.title = element_text(size = 20))


c5 <- bone %>% 
  ggplot(aes(x = survival_status, y = as.numeric(CD3_x1e8_per_kg))) +
  geom_boxplot(fill = "#990000", col = "black") +
  ggtitle("CD3+ Dosage") +
  xlab("Survival Status") +
  ylab("CD3+ Dosage") +
  theme_wsj()+ theme(axis.title=element_text(size=12),
                     plot.title = element_text(size = 20))


c6 <- bone %>% 
  ggplot(aes(x = survival_status, y = as.numeric(CD3_to_CD34_ratio))) +
  geom_boxplot(fill = "#990000", col = "black") +
  scale_y_continuous(trans = "log", breaks = c(2,8,50,100)) +
  ggtitle("CD3+:CD34+ Ratio") +
  xlab("Survival Status") +
  ylab("CD3+:CD34+ Ratio (log scale)") +
  theme_wsj()+ theme(axis.title=element_text(size=12),
                     plot.title = element_text(size = 15))

# plot together
grid.arrange(c1,c2,c3,c4,c5,c6, ncol = 2, nrow = 3)

```

We can observe above the various relationships our variables have with the survival status outcome variable. I can not tell regression outcomes just from graphs, but we are particularly interested in variables such as CD34+ Dosage and CD3+ Dosage which tell us the amount of bone marrow transplanted. 


### Comparison of Outcome (survival_status) and categorical variables

#### Categorical Variables: 
donor_ABO, 

recipient_age_int, 

recipient_ABO, 

disease, 

CMV_status, 

HLA_match..out.of.10., 

antigen, allele, HLA_group_1 

#### Binary Variables:
donor_age_below_35 (yes/no), donor_CMV (present/absent), 
recipient_age_below_10 (yes/no), recipient_rh (plus/minus), 
recipient_CMV (present/absent), disease_group (malignant/nonmalignant), gender_match(female_to_male/other), ABO_match (matched/mismatched), 
HLA_mismatch (matched, mismatched), risk_group (high/low), 
stem_cell_source (peripheral_blood/bone_marrow), tx_post_relapse (yes/no), acute_GvHD_II_III_IV (yes/no), acute_GvHD_III_IV (yes/no), 
extensive_chronic_GvHD (yes/no), relapse (yes/no), 
survival_status (yes,no)

```{r, fig.height=10, fig.width=10}

par(mfrow=c(3,3))

#Donor ABO
#2x2 table with margins
tab1 <- table(bone$survival_status,bone$donor_ABO)
tab1 <-prop.table(tab1,2)
addmargins(tab1)

barplot(tab1,xlab='Donor Blood Type', ylab='Survival = Yellow, Death = Red',
main="Survival by Donor Blood Type", col=c("yellow","#990000"),
legend=rownames(tab1),args.legend=list(x="topleft"), beside=T)



#Recipient age
#2x2 table with margins
tab2 <- table(bone$survival_status,bone$recipient_age_int)
tab2 <-prop.table(tab2,2)
addmargins(tab2)

barplot(tab2,xlab='Recipient Age Range', ylab='Survival = Yellow, Death = Red',
main="Survival by Recipient Age Range", col=c("yellow","#990000"),
legend=rownames(tab2),args.legend=list(x="topleft"), beside=T)



#recipient_ABO
#2x2 table with margins
tab3 <- table(bone$survival_status,bone$recipient_ABO)
tab3 <-prop.table(tab3,2)
addmargins(tab3)

barplot(tab3,xlab='Recipient Blood Type', ylab='Survival = Yellow, Death = Red',
main="Survival by Recipient Blood Type", col=c("yellow","#990000"),
legend=rownames(tab3),args.legend=list(x="topleft"), beside=T)



#Disease
#2x2 table with margins
# disease by survival
tab4 <- table(bone$survival_status,bone$disease)
tab4 <-prop.table(tab4,2)
addmargins(tab4)


barplot(tab4,xlab='Recipient Disease', ylab='Survival = Yellow, Death = Red',
main="Survival by Recipient Disease", col=c("yellow","#990000"),
legend=rownames(tab4),args.legend=list(x="topleft"), beside=T)



#CMV_status

#2x2 table with margins
# CMV status by survival
tab5 <- table(bone$survival_status,bone$CMV_status)
tab5 <-prop.table(tab5,2)
addmargins(tab5)

barplot(tab5,xlab='CMV Status', ylab='Survival = Yellow, Death = Red',
main="Survival by Serological compatibility between donor and the recipient", col=c("yellow","#990000"),
legend=rownames(tab5),args.legend=list(x="topleft"), beside=T)



#HLA Match

#2x2 table with margins
# CMV status by survival
tab6 <- table(bone$survival_status,bone$HLA_match..out.of.10.)
tab6 <-prop.table(tab6,2)
addmargins(tab6)

barplot(tab6,xlab='HLA Match (out of 10)', ylab='Survival = Yellow, Death = Red',
main="Survival by HLA Match", col=c("yellow","#990000"),
legend=rownames(tab6),args.legend=list(x="topleft"), beside=T)




#antigen - 

#2x2 table with margins
# CMV status by survival
tab7 <- table(bone$survival_status,bone$antigen)
tab7 <- prop.table(tab7,2)
addmargins(tab7)

barplot(tab7,xlab='Antigen Difference', ylab='Survival = Yellow, Death = Red',
main="Survival by Antigen Difference between Donor and Recipient", col=c("yellow","#990000"),
legend=rownames(tab7),args.legend=list(x="topleft"), beside=T)



#allele
#2x2 table with margins
# CMV status by survival
tab8 <- table(bone$survival_status,bone$allele)
tab8 <- prop.table(tab8,2)
addmargins(tab8)


barplot(tab8,xlab='Allele Difference', ylab='Survival = Yellow, Death = Red',
main="Survival by Allele Difference between Donor and Recipient", col=c("yellow","#990000"),
legend=rownames(tab8),args.legend=list(x="topleft"), beside=T)



#Need to fix axis labels ------------------

#HLA_group_1

#2x2 table with margins
# CMV status by survival
tab9 <- table(bone$survival_status,bone$HLA_group_1)
tab9 <- prop.table(tab9,2)
addmargins(tab9)


barplot(tab9,xlab='Difference Type', ylab='Survival = Yellow, Death = Red', main="Survival by Difference Type between Donor and Recipient", col=c("yellow","#990000"),
legend=rownames(tab9),args.legend=list(x="topleft"), beside=T)


summary(bone$PLT_recovery)
summary(bone$ANC_recovery)
```


Looking at survival status looking at variable with more than 2 categories:

donor_ABO - Seems that donor of blood type A gives the least likely chances of survival (but not by a considerable amount), but blood donors with type AB show drasticly better chances of survival
recipient_age_int - Recipient ages between 0-5 and increased chances of survival
recipient_ABO - Biggest difference of survival success is a recipient blood type of AB
disease - all disease categories show the same probability of survival except lukemia, which shows a 0% chance of survival
CMV_status - Shows serological compatibility of the donor and the recipient of hematopoietic stem cells according to cytomegalovirus infection prior to transplantation (the higher the value the lower the compatibility). Here we see that lower values like 0 and 1 show a higher chance of survival which is what we expect
HLA_match..out.of.10. - Compatibility of antigens of the main histocompatibility complex of the donor and the recipient of hematopoietic stem cells (10/10, 9/10, 8/10, 7/10) - a compatibility of 7/10 shows the least likely chances of survival
antigen - In how many antigens there is a difference between the donor and the recipient (0-3). An antigen difference of 3 increased the chances of death, while a difference of 2 had a better chance of survival than an antigen different of 1... interesting finding
allele -In how many allele there is a difference between the donor and the recipient (0-4). Pretty much the same chances of survival except when allele different was 4 then there was no chance of survival
HLA_group_1 - The difference type between the donor and the recipient (HLA matched, one antigen, one allele, DRB1 cell, two allele or allel+antigen, two antigenes+allel, mismatched). The DRB1_cell difference had an increased chance of death along with mismatched. Weird finding is that 3 differences increased survival drastically


```{r, fig.height=10, fig.width=10}
par(mfrow = c(4,4))

#Donor Age Group
#2x2 table with margins
tab10 <- table(bone$survival_status,bone$donor_age_below_35)
tab10 <- prop.table(tab10,2)
addmargins(tab10)

barplot(tab10,xlab='Donor Age Under 35', ylab='Survival = Yellow, Death = Red',
main="Survival by Donor Age Range", col=c("yellow","#990000"),
legend=rownames(tab10),args.legend=list(x="bottomleft"), beside=T)


#Donor CMV
#2x2 table with margins
tab11 <- table(bone$survival_status,bone$donor_CMV)
tab11 <- prop.table(tab11,2)
addmargins(tab11)

barplot(tab11,xlab='Donor CMV Status', ylab='Survival = Yellow, Death = Red',
main="Survival by Donor CMV Status", col=c("yellow","#990000"),
legend=rownames(tab11),args.legend=list(x="bottomleft"), beside=T)


#recipient_age_below_10
#2x2 table with margins
tab12 <- table(bone$survival_status,bone$recipient_age_below_10)
tab12 <- prop.table(tab12,2)
addmargins(tab12)

barplot(tab12,xlab='Recipient Age Below 10', ylab='Survival = Yellow, Death = Red',
main="Survival by Recipient Age Range", col=c("yellow","#990000"),
legend=rownames(tab12),args.legend=list(x="bottomleft"), beside=T)

#recipient_rh
#2x2 table with margins
tab13 <- table(bone$survival_status,bone$recipient_rh)
tab13 <- prop.table(tab13,2)
addmargins(tab13)

barplot(tab13,xlab='Recipient Rh Presence', ylab='Survival = Yellow, Death = Red',
main="Survival by Recipient rh Presence", col=c("yellow","#990000"),
legend=rownames(tab13),args.legend=list(x="bottomleft"), beside=T)


#recipient_CMV
#2x2 table with margins
tab14 <- table(bone$survival_status,bone$recipient_CMV)
tab14 <- prop.table(tab14,2)
addmargins(tab14)

barplot(tab14,xlab='Recipient CMV Presence', ylab='Survival = Yellow, Death = Red',
main="Survival by Recipient CMV Presence", col=c("yellow","#990000"),
legend=rownames(tab14),args.legend=list(x="bottomleft"), beside=T)


#disease_group
#2x2 table with margins
tab15 <- table(bone$survival_status,bone$disease_group)
tab15 <- prop.table(tab15,2)
addmargins(tab15)

barplot(tab15,xlab='Disease Group', ylab='Survival = Yellow, Death = Red',
main="Survival by Disease Group", col=c("yellow","#990000"),
legend=rownames(tab15),args.legend=list(x="bottomleft"), beside=T)


#gender_match ***------------------- don't understand this variable
#2x2 table with margins
tab16 <- table(bone$survival_status,bone$gender_match)
tab16 <- prop.table(tab16,2)
addmargins(tab16)

barplot(tab16,xlab='Gender Match', ylab='Survival = Yellow, Death = Red',
main="Survival by Gender Match?", col=c("yellow","#990000"),
legend=rownames(tab16),args.legend=list(x="bottomleft"), beside=T)


#ABO_match
#2x2 table with margins
tab17 <- table(bone$survival_status,bone$ABO_match)
tab17 <- prop.table(tab17,2)
addmargins(tab17)

barplot(tab17,xlab='Match based on Blood Group', ylab='Survival = Yellow, Death = Red',
main="Survival by Match of Blood Group", col=c("yellow","#990000"),
legend=rownames(tab17),args.legend=list(x="bottomleft"), beside=T)



#HLA_match
#2x2 table with margins
tab18 <- table(bone$survival_status,bone$HLA_mismatch)
tab18 <- prop.table(tab18,2)
addmargins(tab18)

barplot(tab18,xlab='Match based on HLA', ylab='Survival = Yellow, Death = Red',
main="Survival by Match of HLA", col=c("yellow","#990000"),
legend=rownames(tab18),args.legend=list(x="bottomleft"), beside=T)



#risk_group
#2x2 table with margins
tab19 <- table(bone$survival_status,bone$risk_group)
tab19 <- prop.table(tab19,2)
addmargins(tab19)

barplot(tab19,xlab='Risk Group', ylab='Survival = Yellow, Death = Red',
main="Survival by Risk Group", col=c("yellow","#990000"),
legend=rownames(tab19),args.legend=list(x="bottomleft"), beside=T)


#stem_cell_source
#2x2 table with margins
tab20 <- table(bone$survival_status,bone$stem_cell_source)
tab20 <- prop.table(tab20,2)
addmargins(tab20)

barplot(tab20,xlab='Stem Cell Source', ylab='Survival = Yellow, Death = Red',
main="Survival by Stem Cell Source", col=c("yellow","#990000"),
legend=rownames(tab20),args.legend=list(x="bottomleft"), beside=T)



#tx_post_relapse
#2x2 table with margins
tab21 <- table(bone$survival_status,bone$tx_post_relapse)
tab21 <- prop.table(tab21,2)
addmargins(tab21)

barplot(tab21,xlab='Relapse After Treatment', ylab='Survival = Yellow, Death = Red',
main="Survival by Relapse after Treatment", col=c("yellow","#990000"),
legend=rownames(tab21),args.legend=list(x="bottomleft"), beside=T)



#acute_GvHD_II_III_IV
#2x2 table with margins
tab22 <- table(bone$survival_status,bone$acute_GvHD_II_III_IV)
tab22 <- prop.table(tab22,2)
addmargins(tab22)

barplot(tab22,xlab='Development of Acute GVHD stage II, III or IV ', ylab='Survival = Yellow, Death = Red',
main="Survival by Development of Acute GVHD (Stage II, III or IV)", col=c("yellow","#990000"),
legend=rownames(tab22),args.legend=list(x="bottomleft"), beside=T)



#acute_GvHD_III_IV
#2x2 table with margins
tab23 <- table(bone$survival_status,bone$acute_GvHD_III_IV)
tab23 <- prop.table(tab22,2)
addmargins(tab23)

barplot(tab23,xlab='Development of Acute GVHD stage III or IV ', ylab='Survival = Yellow, Death = Red',
main="Survival by Development of Acute GVHD (Stage III or IV)", col=c("yellow","#990000"),
legend=rownames(tab23),args.legend=list(x="bottomleft"), beside=T)



#extensive_chronic_GvHD
#2x2 table with margins
tab24 <- table(bone$survival_status,bone$extensive_chronic_GvHD)
tab24 <- prop.table(tab24,2)
addmargins(tab24)

barplot(tab24,xlab='Development of Chronic Acute GVHD', ylab='Survival = Yellow, Death = Red',
main="Survival by Development of Chronic Acute GVHD", col=c("yellow","#990000"),
legend=rownames(tab24),args.legend=list(x="bottomleft"), beside=T)



#relapse
#2x2 table with margins
tab25 <- table(bone$survival_status,bone$relapse)
tab25 <- prop.table(tab25,2)
addmargins(tab25)

barplot(tab25,xlab='Relapse', ylab='Survival = Yellow, Death = Red',
main="Survival by Relapse", col=c("yellow","#990000"),
legend=rownames(tab25),args.legend=list(x="bottomleft"), beside=T)

```

Looking at survival status with binary variables

donor_age_below_35 (yes/no) - if donor was under the age of 35 there was a more likely chance of survival
donor_CMV (present/absent) - not a very large association with survival
recipient_age_below_10 (yes/no) - recipients below 10 had a better chance of survival
recipient_rh (plus/minus) - having rh minus indicated better chances of survival
recipient_CMV (present/absent) - not a very large assocation with survival
disease_group (malignant/nonmalignant) - survival chances increased if disease was nonmalignant
gender_match (female_to_male/other) - no difference to survival chances
ABO_match (matched/mismatched) - a mismatch by blood group increased chances of survival? Interesting finding - would not have expected this
HLA_mismatch (matched, mismatched) - not much a difference in survival outcome
risk_group (high/low) - lower risk group has a better chance of survival (expected)
stem_cell_source (peripheral_blood/bone_marrow) - stem cell source from peripheral blood shows a much better chance for survival
tx_post_relapse (yes/no) - no post treatment relapse indicates around equal chances for survival and death, but if post treatment relapse does occur chances of survival decrease considerably
acute_GvHD_II_III_IV (yes/no) - not a huge difference in survival chances if acute GVHD is present in stage II, III or IV
acute_GvHD_III_IV (yes/no) - same for this variable
extensive_chronic_GvHD (yes/no) - development of chronic acute GVHD increases chances of death drastically
relapse (yes/no) - as we figured, occurence of relapse decrease survival chances drastically
