---
title: "BST 210 Project: Survival Analysis"
author: "Lauren Mock"
date: "11/30/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(survival)
library(survminer)
library(pander)
```


# Stem Cell Source

## Read in Data

```{r, warning=FALSE}
bone <- read.csv("data.csv")

# remove HLA_match_raw (was already converted into a different column)
bone <- select(bone, -HLA_match_raw)


# convert numeric columns to numeric
bone_num <- c("donor_age", "recipient_age", "recipient_body_mass", "CD34_x1e6_per_kg...CD34kgx10d6",
           "CD3_x1e8_per_kg", "CD3_to_CD34_ratio", "ANC_recovery", "PLT_recovery",
           "time_to_acute_GvHD_III_IV", "survival_time", "survival_status")

bone[,bone_num] <- lapply(bone_num, function(x) as.numeric(bone[[x]]))
```


## Compare Kaplan-Meier curves for the two stem cell sources

```{r}
# survival object
survival.obj <- Surv(time = bone$survival_time, event = bone$survival_status)

# overall KM
KM.fit1 <- survfit(survival.obj ~ 1, data = bone)

# plot
plot(KM.fit1, xlab = "Days after Transplant", ylab = "Survival Probability", conf.int = FALSE,
mark.time = TRUE, main = "Kaplan-Meier Survival Curve", lwd = 2)



# KM by treatment
KM.fit2 <- survfit(survival.obj ~ stem_cell_source, data = bone)

# plot
#png("KM_source.png")
plot(KM.fit2, xlab = "Days after Transplant", ylab = "Survival Probability", 
     mark.time = TRUE, conf.int=, col = c("coral", "dodgerblue"), 
     main = "Kaplan-Meier Curves by Stem Cell Source", lwd = 2)
legend(x = 1800, y = 1.05, 
       legend = c("Peripheral Blood", "Bone Marrow"), 
       col = c("dodgerblue", "coral"), 
       bty = "n",
       lty = 1:1,
       lwd = 2,
       cex = 1.3)
#dev.off()
```

**note: 5 people (out of 187) missing CD3**


### log-rank test to compare these two curves

```{r}
survdiff(Surv(bone$survival_time, bone$survival_status) ~ stem_cell_source, data=bone)
```

p-value is 0.05--almost statistically significantly different



# CD3+ dosage

## Compare Kaplan-Meier curves for high vs. low dosage

```{r}
# add new binary columns based on previous research
bone$CD3_over_4 <- ifelse(bone$CD3_x1e8_per_kg >= 4, 1, 0)
bone$CD34_over10 <- ifelse(bone$CD34_x1e6_per_kg...CD34kgx10d6 >= 10, 1, 0)

# model fit
KM.fit4 <- survfit(survival.obj ~ CD3_over_4, data = bone)

# plot
png("KM_CD3.png") 
plot(KM.fit4, xlab = "Days after Transplant", ylab = "Survival Probability", mark.time = TRUE, conf.int=, 
     col = c("coral", "dodgerblue"), lwd = 2,
     main = "Kaplan-Meier Curves by CD3+ Dosage per kg Body Weight")
legend(x = 500, y = 1.05, 
       legend = c(expression(paste("CD3+ dosage >= 4*", 10^8, " per kg weight")), 
                  expression(paste("CD3+ dosage < 4*", 10^8, " per kg weight"))), 
       col = c("dodgerblue", "coral"), 
       bty = "n",
       lty = 1:1,
       lwd = 2,
       cex = 1.3)
dev.off()
```

### log-rank test to compare these two curves

```{r}
survdiff(Surv(bone$survival_time, bone$survival_status) ~ CD3_over_4, data=bone)
```

highly significant difference! (p-value 0.01)


# Cox Proportional Hazards Model

## unadjusted

```{r}
cox_unadj <- coxph(survival.obj ~ as.factor(stem_cell_source) + CD3_x1e8_per_kg, 
                        data = bone, ties = "exact")
summary(cox_unadj)
```

## Confounders?

**looking for common causes of stem cell type and survival or common causes of dosage and survival**

did the authors adjust for confounders?
no, it was an RCT


possible confounders
- always must adjust for age when adjusting for dosage
- disease type (Anasetti article says that this could be an effect modifier by interacting with stem cell source)


### look at dosage distribution by source

```{r, warning=FALSE}
bone %>% ggplot(aes(x = stem_cell_source, y = CD3_x1e8_per_kg)) +
        geom_boxplot() + 
        ggtitle("CD3+ Dosage per kg by Stem Cell Source") +
        xlab("Stem Cell Source") +
        ylab(expression(paste("CD3+ x ", 10^8, " dosage per kg weight"))) +
        scale_x_discrete(labels = c("Bone Marrow", "Peripheral Blood"))

# get means
# bone %>% filter(stem_cell_source == "bone_marrow") %>% summary()
# bone %>% filter(stem_cell_source == "peripheral_blood") %>% summary()
```

- people who get bone marrow always get a much smaller dose
- this could definitely be a confounder, so we'll keep both stem cell source and dose in the model
- interpretation will be "given that someone received this source, what was the effect of the dosage?"

### look at dosage distribution by age

```{r}
bone %>% ggplot(aes(x = recipient_age_below_10, y = CD3_x1e8_per_kg)) +
        geom_boxplot() + 
        ggtitle("CD3+ Dosage per kg by Age") +
        xlab("Stem Cell Source") +
        ylab(expression(paste("CD3+ x ", 10^8, " dosage per kg weight"))) +
        scale_x_discrete(labels = c("Older than 10", "Younger than 10"))
```

younger people are more likely to get higher doses (per kg body weight), which makes sense because they are also more likely to get peripheral blood (given in higher doses than bone marrow)

### look at survival by age

```{r}
bone %>% ggplot(aes(x = as.factor(survival_status), y = recipient_age)) +
        geom_boxplot() +
        ggtitle("Survival by Recipient Age") +
        xlab("Survival Status") +
        ylab("Recipient Age") +
        scale_x_discrete(labels = c("Survived", "Died"))
```



### look at source by recipient age

```{r}
bone %>% ggplot(aes(x = stem_cell_source, y = recipient_age)) +
        geom_boxplot() +
        ggtitle("Recipient Age by Stem Cell Source") +
        xlab("Stem Cell Source") +
        ylab("Recipient Age") +
        scale_x_discrete(labels = c("Bone Marrow", "Peripheral Blood"))
```
reasonably similar age distributions

### look at source by donor age

```{r}
bone %>% ggplot(aes(x = stem_cell_source, y = donor_age)) +
        geom_boxplot() +
        ggtitle("Donor Age by Stem Cell Source") +
        xlab("Stem Cell Source") +
        ylab("Recipient Age") +
        scale_x_discrete(labels = c("Bone Marrow", "Peripheral Blood"))
```

reasonably similar age distributions

### GVHD by dose

```{r}
bone %>% ggplot(aes(x = acute_GvHD_II_III_IV, y = CD3_x1e8_per_kg)) +
        geom_boxplot() +
        ggtitle("CD3+ Dosage and Acute GVHD") +
        xlab("Presence of Acute GVHD (Stage II, III, or IV)") +
        ylab(expression(paste("CD3+ x ", 10^8, " dosage per kg weight (log scale)"))) +
        scale_x_discrete(labels = c("No", "Yes")) + 
        scale_y_log10() +
        facet_grid( ~ stem_cell_source, 
                    labeller = as_labeller(c(`bone_marrow` = "Bone Marrow", 
                                             `peripheral_blood` = "Peripheral Blood")))
```

no apparent relationship between high dose of CD3+ and GVHD for either stem cell source

## confounders

### adjusted model 1

```{r}
cox_adj1 <- coxph(survival.obj ~ as.factor(stem_cell_source) + CD3_x1e8_per_kg + recipient_age, 
                        data = bone, ties = "exact")
summary(cox_adj1)
```

adding age changed the stem cell source coefficient by > 10%


### adjusted model 2

check interaction between stem cell source and disease type (Anasetti et al.)

```{r}
cox_adj2 <- coxph(survival.obj ~ as.factor(stem_cell_source):as.factor(disease), 
                    data = bone, ties = "exact")
summary(cox_adj2)
```

definitely some significant results, but need to check that we have enough people in each covariate pattern


#### check covariate patterns for disease type and stem cell source interaction

```{r}
cov_patterns <- bone %>% 
        group_by(disease, stem_cell_source) %>%
        count() %>%
        spread(stem_cell_source, n) %>%
        rename(Disease = "disease",
                `Bone Marrow` = "bone_marrow",
                `Peripheral Blood` = "peripheral_blood")

# rename ALL and AML
cov_patterns$Disease[1:2] <- c("acute lymphoblastic leukemia", "acute myeloid leukemia")

# print table
cov_patterns %>% pander()
```

Peripheral blood was more common than bone marrow for all five diseases. Unfortunately, we do not have enough people in each covariate pattern to include this interaction term in our model. For example, all 9 patients who had lymphoma died, so we cannot look at the interaction between lymphoma and stem cell source.


### adjusted model 3

```{r}
cox_adj3 <- coxph(survival.obj ~ as.factor(stem_cell_source) + CD3_x1e8_per_kg + 
                          recipient_age + as.factor(disease), 
                        data = bone, ties = "exact")
summary(cox_adj3)

# bone %>% 
#         group_by(CMV_status, stem_cell_source) %>%
#         count() %>%
#         pander()
```

adding disease changes coefficient by 11.4%



**final model = adjusted model 3**

## does the benefit of a higher dose of CD3+ just keep increasing?

```{r}
bone %>% ggplot(aes(x = as.factor(survival_status), y = CD3_x1e8_per_kg)) +
        geom_boxplot() +
        ggtitle("CD3+ Dosage and Survival") +
        xlab("Survival Status at End of Follow-Up") +
        ylab(expression(paste("CD3+ x ", 10^8, " dosage per kg weight (log scale)"))) +
        scale_x_discrete(labels = c("Survived", "Died")) + 
        scale_y_log10() +
        facet_grid( ~ stem_cell_source, 
                    labeller = as_labeller(c(`bone_marrow` = "Bone Marrow", 
                                             `peripheral_blood` = "Peripheral Blood")))
```

People who survived seem to have gotten higher CD3+ doses among both stem cell source types (but this may not be statistically significant).



## check Schoenfeld residuals for proportional hazards assumption

```{r}
# create plot of schoenfeld resids
wt_sch_dose <- cox.zph(cox_adj1)
plot(wt_sch_dose) # slight decrease then increase over time

# check summary to see if problematic
wt_sch_dose
```

```{r}

png("both_KM.png", width = 1000, height = 500)

par(mfrow = c(1,2))


plot(KM.fit2, xlab = "Days after Transplant", ylab = "Survival Probability", 
     mark.time = TRUE, conf.int=, col = c("coral", "dodgerblue"), 
     main = "Kaplan-Meier Curves by Stem Cell Source", lwd = 2)
legend(x = 1800, y = 1.05, 
       legend = c("Peripheral Blood", "Bone Marrow"), 
       col = c("dodgerblue", "coral"), 
       bty = "n",
       lty = 1:1,
       lwd = 2,
       cex = 1.3)

plot(KM.fit4, xlab = "Days after Transplant", ylab = "Survival Probability", mark.time = TRUE, conf.int=, 
     col = c("coral", "dodgerblue"), lwd = 2,
     main = "Kaplan-Meier Curves by CD3+ Dosage per kg Body Weight")
legend(x = 500, y = 1.05, 
       legend = c(expression(paste("CD3+ dosage >= 4*", 10^8, " per kg weight")), 
                  expression(paste("CD3+ dosage < 4*", 10^8, " per kg weight"))), 
       col = c("dodgerblue", "coral"), 
       bty = "n",
       lty = 1:1,
       lwd = 2,
       cex = 1.3)

dev.off()
```


